Retrofitting Effect Handlers onto OCaml

### 2020-11-20 retro-concurrency.tex
Updates

Line 4 declares an effect |DivBy0| that is parameterised by an integer, which
when performed returns the result of division-by-zero. If the denominator of
the division is 0, then we perform the |DivBy0| effect with the numerator value
(line 12) so that the context may decide the return value. We obtain the |iOS|
semantics by wrapping the evaluator with this handler:
\begin{lstlisting}
let ios e = match eval e with
| v -> v (* value case *)
| effect (DivBy0 _) k -> continue k Error (* effect case *)
\end{lstlisting}
If the evaluator returns a value, we simply return that. If the evaluator
performs the |DivBy0| effect, we catch the effect in the pattern match along with
the continuation |k| of the corresponding |perform| \emph{delimited} by this
handler. For |iOS|, we |continue| the suspended computation with an
|Error| value and the computation |ios(1/0)| returns |Error|.
We can get the |gsearch| semantics by enclosing the evaluator in this handler:
\begin{lstlisting}
let gsearch e = match eval e with
| v -> v
| effect (DivBy0 v) k ->
		if v = 0 then continue k Error
		else if v > 0 then continue k (Inf P)
		else continue k (Inf N)
\end{lstlisting}
Executing |gsearch(1/0)| returns positive infinity (|Inf P|).

@@

OCaml currently does not support any non-local control flow mechanisms other
than exceptions. This makes it particularly challenging to implement the
delimited continuations necessary for effect handlers without sacrificing the
desirable properties of OCaml. One standard way of implementing continuations is
to use continuation-passing style (CPS) in the compiler's intermediate
representation (IR)~\cite{Leijen14}.
%If the compiler were to use CPS, then
%reifying continuations in the source language is relatively straight-forward.
OCaml does not use a CPS IR, and changing the compiler to utilise a CPS IR
would be an enormous undertaking that would affect the
performance profile of existing OCaml applications due to increased memory allocations as
continuation closures get allocated on the heap~\cite{Farvardin20}. Moreover,
with CPS, stack unwinding is meaningless and hence, we would lose compatibility
with tools that inspect the program stack. Hence, we choose not to use CPS
translation and represent continuations as call stacks.

@@

Our work is also timely. The WebAssembly~\cite{Hass17} community group is
considering effect handlers as one of the mechanisms for supporting
concurrency, asynchronous I/O and generators~\cite{WasmProposal}. Project
Loom~\cite{loom} is an OpenJDK project that adds virtual threads and delimited
continuations to Java. The Swift roadmap~\cite{swift} includes
direct-style asynchronous programming and structured concurrency as milestones.
We believe that our design choices will inform similar choices to be made in the
other industrial-strength languages.

@@

OCaml does not use the C calling convention. In particular, there are no
callee-saved registers in OCaml. In the |x86-64| backend, the OCaml runtime
makes use of two C callee-saved registers for supporting OCaml execution. The
register |r15| holds the \emph{allocation pointer} into the minor heap used for
bump pointer allocation, and |r14| holds a reference to the |Caml_state| table
of global variables used by the runtime. This makes external calls extremely
fast in OCaml. If the external function does not allocate in the OCaml heap,
then it can be called directly and no bookkeeping is necessary. For external
functions which allocate in the OCaml heap, the cached allocation pointer is
saved to |Caml_state| before the external call and it is restored on return.
Similarly, callbacks into OCaml are also cheap: these involve loading the
arguments in the right registers and calling the OCaml function. OCaml
callbacks are relatively common as the garbage collector (GC) finalisation
functions are executed as OCaml callbacks.

@@

\subsubsection{Syntax} Our expressions (Figure~\ref{sem:syntax}) consists of
integer constants ($n$), variables ($x$), abstraction ($\lam{x}{x}$),
application ($e~e$), arithmetic expressions ($e \odot e$) where $\odot$ ranges
over $\{+,-,*,/\}$, raising exceptions ($\throw{l}{e}$), performing effects
($\perform{l}{e}$), and handling effects ($\handle{e}{h}$). Abstractions come
in two forms: OCaml abstractions ($\lambda^o$) and C abstractions
($\lambda^c$). The handler consists of a return case ($\caseval{x}{e}$), zero
or more exception cases ($\caseexn{l}{x}{e}$) with label $l$, parameter $x$ and
body $e$, and zero or more effect cases ($\caseeff{l}{x}{k}{e}$) with label
$l$, parameter $x$, continuation $k$ and body $e$.

@@

In the operational semantics, the continuations may be resumed more than once.
Captured continuations are copies of the original fibers and resuming the
continuation copies the fibers and leaves the continuation as it is. In our use
cases, continuations will be resumed at-most once, and copying fibers is
unnecessary and inefficient. Instead, Multicore OCaml optimises fibers for
one-shot continuations.

Fibers are allocated on the heap using |malloc| and are |free|d when no longer
necessary. Similar to Farvardin et al.~\cite{Farvardin20}, we use a \emph{stack
cache} of recently freed stacks in order to speed up allocation.

@@

@@

\vspace{-2mm}
\subsection{External calls and callbacks}

@@

\vspace{-2mm}
\subsection{Effect handlers}
\label{sec:effimpl}

@@

We also present micro benchmarks results in Table~\ref{tab:micro_noeffect}.
Since micro benchmarks magnify micro-architectural optimisations, we also
report the number of instructions executed (obtained using |perf|) along with
time. |exnval| performs 100 million iterations of installing exception handlers
and returning with a value. |exnraise| is similar, but raises an exception in
each iteration. |extcall| and |callback| perform 100 million external calls and
callbacks to identity functions. The other micro benchmarks are highly
recursive programs and were taken from ~\cite{Farvardin20}. For micro benchmarks,
we observed that padding tight loops with a few |nop| instructions, which
changes the loop alignment, can make the code up to 15\% faster. Hence, the
difference in running times under 15\% may not be statistically significant.

@@

\textbf{\textit{Generators.}} Generators allow data structures to be traversed on
demand. Many languages including JavaScript and Python provide generators as a
built-in primitive. Using effect handlers (|MC|), given any data structure
(|'a t|) and its iterator (|val iter: 'a t -> ('a -> unit) -> unit|), we can
derive its generator function (|val next : unit -> 'a option|). The code is
available in the supplementary material. We evaluate the performance of
traversing a complete binary tree of depth 25 using this generator. This
involves $2^{26}$ stack switches in total. For comparison, we also implemented
a hand-written, selective CPSed~\cite{Nielson01},
defunctionalised~\cite{Danvy01} version (|cps|) and a concurrency monad
(|monad|) version of the generator for the tree. Both |cps| and |monad|
versions are specialised to the binary tree, and come with the usual caveats of
not using the stack for function calls. We observed that the |cps| version was
the fastest thanks to specialisation and hand optimisation. |MC| version was
only 2.76$\times$ slower than |cps| while being a generic
solution, whereas the |monad| version was 8.69$\times$ slower than |cps|.

>>>>

Line 4 declares an effect |DivBy0| that is parameterised by an integer, which
when performed returns the result of division-by-zero. If the denominator of
the division is |0|, then we perform the |DivBy0| effect with the numerator
value (line 12) so that the context may decide the return value. We obtain the
|iOS| semantics by wrapping the evaluator with this handler:
\begin{lstlisting}
let ios e = match eval e with
| v -> v (* value case *)
| effect (DivBy0 _) k -> continue k Error (* effect case *)
\end{lstlisting}
If the evaluator returns a value, we simply return that. If the evaluator
performs the |DivBy0| effect, we catch the effect in the pattern match along with
the continuation |k| of the corresponding |perform| \emph{delimited} by this
handler. For |iOS|, we |continue| the suspended computation with an
|Error| value and the computation |ios(1/0)| returns |Error|.
We can get the |gsearch| semantics by enclosing the evaluator in this handler:
\begin{lstlisting}
let gsearch e = match eval e with
| v -> v
| effect (DivBy0 v) k ->
		if v = 0 then continue k Error
		else if v > 0 then continue k (Inf P)
		else continue k (Inf N)
\end{lstlisting}
Executing |gsearch(1/0)| returns positive infinity (|Inf P|).

@@

OCaml currently does not support any non-local control flow mechanisms other
than exceptions. This makes it particularly challenging to implement the
delimited continuations necessary for effect handlers without sacrificing the
desirable properties of OCaml. One standard way of implementing continuations
is to use continuation-passing style (CPS) in the compiler's intermediate
representation (IR)~\cite{Leijen14}. OCaml does not use a CPS IR, and changing
the compiler to utilise a CPS IR would be an enormous undertaking that would
affect the performance profile of existing OCaml applications due to increased
memory allocations as continuation closures get allocated on the
heap~\cite{Farvardin20}. Moreover, with CPS, stack unwinding is meaningless and
hence, we would lose compatibility with tools that inspect the program stack.
Hence, we choose not to use CPS translation and represent continuations as call
stacks.

@@

Our work is also timely. The WebAssembly~\cite{Hass17} community group is
considering effect handlers as one of the mechanisms for supporting
concurrency, asynchronous I/O and generators~\cite{WasmProposal}. Project
Loom~\cite{loom} is an OpenJDK project that adds virtual threads and delimited
continuations to Java. The Swift roadmap~\cite{swift} includes direct-style
asynchronous programming and structured concurrency as milestones. We believe
that our design choices will inform similar choices to be made in other
industrial-strength languages.

@@

OCaml does not use the C calling convention. In particular, there are no
callee-saved registers in OCaml. In the |x86-64| backend, the OCaml runtime
makes use of two C callee-saved registers for supporting OCaml execution. The
register |r15| holds the \emph{allocation pointer} into the minor heap used for
bump pointer allocation, and |r14| holds a reference to the |Caml_state|, a
table of global variables used by the runtime. This makes external calls
extremely fast in OCaml. If the external function does not allocate in the
OCaml heap, then it can be called directly and no bookkeeping is necessary. For
external functions which allocate in the OCaml heap, the cached allocation
pointer is saved to |Caml_state| before the external call and it is restored on
return. Similarly, callbacks into OCaml are also cheap: these involve loading
the arguments in the right registers and calling the OCaml function. OCaml
callbacks are relatively common as the garbage collector (GC) finalisation
functions are executed as OCaml callbacks.

@@

\subsubsection{Syntax} Our expressions (Figure~\ref{sem:syntax}) consist of
integer constants ($n$), variables ($x$), abstraction ($\lam{x}{x}$),
application ($e~e$), arithmetic expressions ($e \odot e$) where $\odot$ ranges
over $\{+,-,*,/\}$, raising exceptions ($\throw{l}{e}$), performing effects
($\perform{l}{e}$), and handling effects ($\handle{e}{h}$). Abstractions come
in two forms: OCaml abstractions ($\lambda^o$) and C abstractions
($\lambda^c$). The handler consists of a return case ($\caseval{x}{e}$), zero
or more exception cases ($\caseexn{l}{x}{e}$) with label $l$, parameter $x$ and
body $e$, and zero or more effect cases ($\caseeff{l}{x}{k}{e}$) with label
$l$, parameter $x$, continuation $k$ and body $e$.

@@

In the operational semantics, the continuations may be resumed more than once.
Captured continuations are copies of the original fibers and resuming the
continuation copies the fibers and leaves the continuation as it is. In our use
cases, continuations will be resumed at-most once, and copying fibers is
unnecessary and inefficient. Instead, Multicore OCaml optimises fibers for
one-shot continuations. Fibers are allocated on the heap using |malloc| and are
|free|d when no longer necessary. Similar to Farvardin et
al.~\cite{Farvardin20}, we use a \emph{stack cache} of recently freed stacks in
order to speed up allocation.

@@

%It is useful to note that stack overflow is a rare operation
%and the CPU branch predictor will correctly predict the branch with high
%accuracy.

@@

\vspace{-3mm}
\subsection{External calls and callbacks}

@@

\vspace{-3mm}
\subsection{Effect handlers}
\label{sec:effimpl}

@@

We also present micro benchmarks results in Table~\ref{tab:micro_noeffect}.
Since micro benchmarks magnify micro-architectural optimisations, we also
report the number of instructions executed (obtained using |perf|) along with
time. |exnval| performs 100 million iterations of installing exception handlers
and returning with a value. |exnraise| is similar, but raises an exception in
each iteration. |extcall| and |callback| perform 100 million external calls and
callbacks to identity functions. The other micro benchmarks are highly
recursive programs and were taken from ~\cite{Farvardin20}. For micro
benchmarks, we observed that padding tight loops with a few |nop| instructions,
which changes the loop alignment, makes the code up to 15\% faster. Hence, the
difference in running times under 15\% may not be statistically significant.

@@

\textbf{\textit{Generators.}} Generators allow data structures to be traversed on
demand. Many languages including JavaScript and Python provide generators as a
built-in primitive. Using effect handlers (|MC|), given any data structure
(|'a t|) and its iterator (|val iter: 'a t -> ('a -> unit) -> unit|), we can
derive its generator function (|val next : unit -> 'a option|). The code is
available in the supplementary material. We evaluate the performance of
traversing a complete binary tree of depth 25 using this generator. This
involves $2^{26}$ stack switches in total. For comparison, we also implemented
a hand-written, selective CPSed~\cite{Nielson01},
defunctionalised~\cite{Danvy01} version (|cps|) and a concurrency monad
(|monad|) version of the generator for the tree. Both |cps| and |monad|
versions are specialised to the binary tree, and come with the usual caveats of
not using the stack for function calls. We observed that the |cps| version was
the fastest, thanks to specialisation and hand optimisation. |MC| version was
only 2.76$\times$ slower than |cps| while being a generic
solution, whereas the |monad| version was 8.69$\times$ slower than |cps|.
